Rediscovering the Rete Ovarii: a secreting auxiliary structure to the ovary

Version 1

Abstract

Abstract

The rete ovarii (RO) is an appendage of the ovary that has been given little attention. Although the RO appears in drawings of the ovary in early versions of Gray’s Anatomy, it disappeared from recent textbooks, and is often dismissed as a functionless vestige in the adult ovary. Using PAX8 immunostaining and confocal microscopy, we characterized the fetal development of the RO in the context of the ovary. The RO consists of three distinct regions that persist in adult life, the intraovarian rete (IOR), the extraovarian rete (EOR), and the connecting rete (CR). While the cells of the IOR appear to form solid cords within the ovary, the EOR rapidly develops into a convoluted tubular epithelium ending in a distal dilated tip. Cells of the EOR are ciliated and exhibit cellular trafficking capabilities. The CR, connecting the EOR to the IOR, gradually acquires tubular epithelial characteristics by birth. Using microinjections into the distal dilated tip of the EOR, we found that luminal contents flow towards the ovary. Mass spectrometry revealed that the EOR lumen contains secreted proteins potentially important for ovarian function. We show that the cells of the EOR are closely associated with vasculature and macrophages, and are contacted by neuronal projections, consistent with a role as a sensory appendage of the ovary. The direct proximity of the RO to the ovary and its integration with the extraovarian landscape suggest that it plays an important role in ovary development and homeostasis.

Introduction

The Rete Ovarii (RO) is an epithelial structure that is directly connected to the ovary, first described over a century ago (Waldeyer, 1870) as a multi-region structure of mesonephric origin. Despite high conservation among mammalian species including guinea pigs, cows, cats, sheep, swine, prairie deer mice, camels, dogs, monkeys (Cassali et al., 2000; Kim et al., 2012), cows (Santos et al., 2012), and humans (Khan et al., 1999), previous work on the RO did not arrive at a consensus on the function of this structure. Thus, it has remained a mysterious orphan structure.

Although the RO has been proposed to be the female homologue of the rete testis (Wenzel and Odend’hal, 1985), our recent work revealed that it is more complex (McKey et al., 2022). The RO is composed of three distinct regions: (1) the extraovarian rete (EOR) which consists of columnar epithelial cells that create a single convoluted tubule structure ending in a blind distal dilated tip (DDT), (2) the connecting rete (CR) which consists of pseudo-columnar cells; and (3) the intraovarian rete (IOR) which consists of squamous epithelial cells that branch and form a fine network of thin solid cell cords approximately 1-3 cells thick (Byskov and Lintern-Moore, 1973; McKey et al., 2022). It was previously shown that the RO consists of ciliated and non-ciliated cells rich in apical microvilli and mitochondria (Czernobilsky et al., 1985). Fine granular Periodic acid-Schiff (PAS) positive material was found in the cytoplasm of the EOR and CR cells, indicating the presence of polysaccharides (e.g. glycogen and mucins) (Stein and Anderson, 1979). Because the proportion of PAS-positive material in the RO was found to be influenced by the estrous cycle in cows (Wenzel et al., 1987), researchers concluded that the luminal contents of the RO were under endocrine control (Archbald et al., 1971). Although these results strongly suggested a secretory role for the RO, this function was not experimentally confirmed, and no proteomic or metabolomic investigations were pursued. Previous characterization of the RO relied heavily on imaging serial sections to study the structure. However, these studies lacked contextual information, and the use of sections made it challenging to determine whether tubules were connected to one another or were isolated structures (Wenzel and Odend’hal, 1985; Woolnough et al., 2000). Furthermore, some investigators reported that the whole RO contracted postnatally and that the DDT of the EOR separated from the RO and degenerated (Byskov and Lintern-Moore, 1973). Perhaps for these reasons, the RO was deemed a functionless vestige, and has been omitted from recent textbook representations of the female reproductive tract (Girsh, 2021).

The RO was rediscovered and highlighted in our recent study (McKey et al., 2022), where we used confocal and lightsheet imaging of whole ovaries to study the integration of ovary morphogenesis with the development of surrounding tissues, including the RO. Recently, the IOR has gained attention as a newly described progenitor for supporting cells of the murine gonad (Mayère et al., 2022). Recent single cell transcriptomics studies have also identified cells of the RO within human fetal gonads (Lardenois et al., 2023; Taelman et al., 2022). We reported that the entire RO expresses high levels of PAX8, and used this as a marker to visualize and characterize cells of the RO. In the present study, we used the Pax8-rtTa; Tre-H2B-Gfp RO nuclear reporter mouse line, advanced imaging techniques, and secretome analysis to characterize the development of the intact RO in its native context and to investigate the function and heterogeneity of its cells. We found that the RO arises from a subset of the mesonephric tubules, analogous to the rete testis, and persists into adulthood. Our studies reveal that the RO is a continuous structure, surrounded by smooth muscle actin, a dense vascular network, and several macrophage populations. We also show that the RO is directly contacted by neurons. The enrichment of secretory machinery in RO cells as well as our experimental analysis of directional flow and luminal contents, together suggest that the RO sends material to the ovary. Based on these findings, we suggest that the RO plays a role in ovary function and should be investigated as a functional organ of the female reproductive tract.

Results

Characterization of the IOR, CR, and EOR

We used immunofluorescence (IF) and confocal imaging to investigate the development and sub -regional structure of the RO from embryonic day (E) 16.5 to postnatal day (P) 7. Three distinct regions of the RO were originally defined histologically by cell morphology (Byskov and Lintern-Moore, 1973; Lee et al., 2011). Using IF and confocal imaging, we found that these three regions are maintained throughout development, but their relative sizes change. While the entirety of the RO is PAX8+, we took different approaches to identify region specific markers. First, as GFRa1 was known to be expressed in the rete testis, we hypothesized that it was expressed in the RO as well. Indeed, immunostaining using antibodies against GFRa1 revealed that part of the RO is GFRa1+. GFRa1 specifically labeled the CR from E16.5 to P7 (Fig. 1 bottom row). To identify additional RO markers, we performed bulk and single cell RNA sequencing (ScRNA-seq) of cells from the ovarian complex of mice at E16.5 and 2 months and found that Krt8 was enriched in the RO (Anbarci et al., 2023). We validated these findings using IF against KRT8, which revealed that KRT8+ cells were specifically localized to the EOR (Fig. 1 third row).

The Rete Ovarii undergoes dynamic changes during development alongside the ovary.

(a,b,c,d top row) Diagrams depicting the development of the three regions (EOR-magenta; CR-cyan; IOR-yellow) of the RO from E16.5-P7 in the whole ovarian complex (ovary-light pink; oviduct-grey). (a,b,c,d bottom rows) Maximum intensity projection from confocal Z-stacks of whole ovary/mesonephros complexes at E16.5 (a), E18.5 (b), P0 (c) and P7 (d) immunostained for PAX8 (yellow), GFRa1 (cyan) and KRT8 (magenta). a,b,c,d second row are composite images, while the third row show separate panels for KRT8 and and the bottom row show panels for GFRa1. *Note that GFRa1 and KRT8 do not co-localize and are specific to the CR and EOR, respectively. All figures are dorsal views of the ovary. Yellow asterisk indicates opening of the infundibulum for reference. Scale bar – 100um

To characterize RO development, we next analyzed PAX8 and KRT8 expression from E16.5 to P0. At E16.5, the CR and IOR were the largest of the three regions. In the EOR, several tubules leading from the region of the CR converged into a single tube of columnar PAX8+ and KRT8+ epithelial cells that lead to a blind end. Between E16.5 and E18.5, the EOR underwent rapid expansion, and the blind end of the EOR became dilated. This structure, which we refer to as the distal dilated tip (DDT) was best visualized from the ventral side of the ovary (SFig. 1). By E18.5, the EOR was the largest of the three regions. At this stage the CR was still large, but the IOR began to regress. At P0, the EOR remained the largest region, and the IOR had regressed to the medullary region of the ovary (McKey et al., 2022). Their distinct structure and protein expression patterns suggest that each region has a different function.

Integration of the EOR with the extra-ovarian environment

During gonadogenesis, the mesonephros is highly vascularized in both XX and XY embryos (Cool et al., 2011). In contrast, the ovary and surrounding tissue are more highly innervated than the testis (McKey et al., 2019). A surprising finding in our bulk transcriptome analysis of the RO was the presence of a high proportion of immune cells and cells with vascular markers that co-isolated with E16.5 RO cells (Anbarci et al., 2023). To explore the integration of the RO with its environment, we used IF to investigate the expression of endothelial marker Endomucin (Fig. 2a), smooth muscle marker alpha smooth muscle actin (aSMA) (Fig. 2b), pan-neuronal marker TUJ1 (Fig. 2c), and macrophage markers F4/80 and LYVE1 (Fig. 2d). We found that at E18.5 the EOR was tightly surrounded by vasculature and ensheathed within a layer of aSMA+ mesenchyme. The EOR, and more specifically the DDT, was directly innervated by neurons that contact the PAX8+ epithelial cells. We also found that the EOR is specifically associated with F4/80+ macrophages (Fig. 4d). This multifaceted integration with the environment suggests the RO may respond to or interpret homeostatic cues.

The EOR is highly integrated with its extra-ovarian environment suggesting multifaceted communication.

(a) Top panel is a whole ovarian complex maximum intensity projection of the confocal Z-stack at E18.5 immunostained for ENDOMUCIN (magenta) and PAX8 (cyan). Bottom panel is an optical section showing vasculature tightly surrounding the DDT of the EOR (magenta). (b) Top panel is a whole ovarian complex maximum intensity projection of the confocal Z-stack at E18.5 immunostained for aSMA (magenta) and KRT8 (cyan). Bottom panel is an optical section showing the EOR tightly ensheathed by smooth muscle (magenta). (c) Top panel is a whole ovarian complex maximum intensity projection of the confocal Z-stack at E18.5 immunostained for TUJ1 (magenta) and PAX8 (cyan). Bottom panel is an optical section showing direct contacts between the EOR and neuronal projections (magenta). (d) EOR at P0 immunostained for PAX8 (yellow), F4/80 (magenta) and LYVE-1 (cyan). Top and bottom panel are maximum intensity projection of the confocal Z-stack. Bottom panel shows the absence of LYVE1 macrophages (cyan) proximate to the EOR, in contrast to the closely associated F4/80 macrophages (magenta in top image). Arrowheads show regions devoid of LYVE1 macrophages where F4/80 macrophages are present. Scale bar – 100um

Flow of luminal material within the EOR

Because the EOR lumen is fluid filled, and because the EOR and CR cells are rich in proteoglycans (Byskov, 1975) we hypothesized that the EOR produces luminal secretions. To investigate this idea, we first chose to determine the direction of flow at P7, when the DDT of the EOR was fully dilated (Fig. 1d). We first injected fluorescently labeled pH-insensitive Dextran into the DDT of the EOR and found that, within just 15 min, the fluorescent fluid had readily traveled from the DDT into the ovary, where it then diffused widely (Fig. 3). By contrast, when we injected fluorescently labeled dextran into the P7 ovary near the IOR, the dextran remained in the ovary and did not travel to the EOR (Fig. 3). These data indicate that, at least at P7, the fluid inside the lumen of the EOR travels towards the ovary.

Fluid moves from the EOR to the ovary.

(a,b) Schematic of dextran injections (syringe) for each group (a-EOR injections, b-ovary injections) (Oviduct-grey, ovary-light pink, IOR-yellow, CR-cyan, EOR-magenta). (Second row) Maximum intensity projection of a confocal Z-stack of whole ovarian complexes at P7, where Dextran was injected into the EOR. Presence of Dextran in the ovary shows that when dextran is injected into the EOR, it diffuses throughout the ovary (dextran-cyan, E-CADHERIN-magenta). (bottom row) Maximum intensity projection from confocal Z-stacks of whole ovary/mesonephros complexes at P7 where Dextran was injected into the ovary. Absence of Dextran signal in the EOR shows that dextran did not travel into the EOR when injected into the ovary near the IOR. Scale bar – 100um

Potential of the EOR and CR for fluid transfer

Without the obvious indication of a lumen within the CR at P7 and taking into consideration the absence of the epithelial marker KRT8 in those cells, we wondered how fluid could travel through the CR to the ovary. Using an antibody against E-CADHERIN, a marker of cell junctions, we found that it was specifically present in the EOR at E16.5, but absent from the CR (Fig. 4a, second row). However, E-Cadherin expression was gradually gained in the CR and IOR such that, by P7, the entire RO was positive for this epithelial marker (Fig. 4a-d, second row). These results suggest that tubular epithelial connections between the EOR and IOR mature gradually between E16.5-P7.

The EOR and CR acquire the potential for fluid transfer

(a,b,c,d) Cells of the CR acquire E-CADHERIN soon after birth. Maximum intensity projection from confocal Z-stacks of whole ovary/mesonephros complexes at E16.5-P7 immunostained for PAX8 (magenta) and E-CADHERIN (cyan). Arrowhead shows the region of the CR where E-CADHERIN expression is negligible at E16.5 (a), low at E18.5 (b) and fully expressed at P0 (c) and P7 (d). (e) Optical section from confocal Z-stacks of RO complexes at P0 immunostained for E-CADHERIN (yellow), CNN1 (magenta), and aSMA (cyan). Boxed regions show the absence of CNN1 in the aSMA+ sheath around the EOR at P0, and (f) its acquisition by P7, suggesting gain of contractility. (g) The RO expresses cilia marker ARL13b by P0. Maximum intensity projection from confocal Z-stacks of RO complexes at E18.5 immunostained for E-CADHERIN (cyan) and ARL13b (magenta). Outlined boxes show all regions of the EOR have ciliated cells. Scale bar – 100um (h). Airyscan optical section from confocal Z-stacks of RO complexes at E18.5 immunostained for KRT8 (cyan) and ARL13b (magenta). Scale bar – 25um

Next, we investigated mechanisms that could facilitate fluid movement through the RO. Previous reports showed that ciliated and non-ciliated cells are present in the RO (Lee et al., 2011). Using an antibody against a marker for cilia, ARL13b, we found that primary ciliated cells were abundant in the DDT and throughout the tubules of the EOR at E18.5 (Fig. 4g-h). Due to the curvature of the DDT, optical slices often include cilia from neighboring cells. However, an ultrathin (0.6um) optical section suggested that each cell has a single cilium. Since the EOR was covered in a sheath of aSMA+ mesenchyme (Fig. 2b, Fig. 4e-f), we investigated whether and when this layer of smooth muscle became contractile. We used an antibody against the contractile smooth muscle protein Calponin (CNN1) and found that CNN1 was absent around the EOR during fetal development and in neonates. However, by P7, the mesenchymal sheath around the EOR gained expression of CNN1, indicating that it acquired the ability to contract by this stage of development (Fig. 4e-f). These data suggest that ciliary mechanosensing and/or muscle contraction may aid the directional movement of the fluid from the DDT to the ovary.

Proteins produced by the EOR indicate a role for the SNARE-complex

Our data showing fluid flow from the EOR to the ovary prompted us to investigate the nature and identity of the proteins produced and potentially secreted by the EOR. To address this question, we analyzed the protein contents of the luminal material using mass spectrometry. Using the Pax8rtTA; Tre-H2B-GFP reporter mouse line, we dissected EORs and isolated luminal fluid by gently pressing the tissue with a pestle. Using this method, we anticipated several problems. First, we expected that some cells within the EOR and surrounding tissue would be lysed during this procedure, which would release proteins not ordinarily secreted. Second, we expected that some cells closely associated with the EOR would be co-isolated, which could result in contamination with contents from cells that are not part of the EOR. To eliminate intracellular proteins arising from lysed cells, we cross-referenced our proteomic dataset with that of the mammalian secretome (Meinken et al., 2015), thereby retaining only known secreted proteins. Next, we compared the resulting list to the E16.5 ScRNA-seq data from cells mapping to the RO (Anbarci et al., 2023). This produced a conservative candidate list of 15 proteins (Table 1). Of the candidate proteins, two were selected for validation, CLU and STXBP2, due to their roles in protein and vesicle transport (Argraves and Morales, 2004; Saewu et al., 2017; Söllner, 2003).

Because no validated antibodies were commercially available to visualize protein expression for CLU and STXBP2, validations were performed using Hybridization Chain Reaction (HCR), a method for single - molecule RNA-Fluorescence in situ hybridization (FISH) (https://files.molecularinstruments.com/MI-Protocol-RNAFISH-Mouse-Rev9.pdf). We found that both Clu and Stxbp2 were expressed in the EOR at E18.5 (Fig. 5a-b) and at P7 (Fig. 5c-d). The presence of Stxbp2 showed that components of the SNARE-complex were actively transcribed in the EOR, suggesting this may serve as a mechanism for secretion. The presence of other components of the SNARE-complex were validated using IF. We found that the T-SNARE complex member STX3was expressed throughout the EOR and was localized to the apical surface, a cellular position that is typically associated with active secretion (Fig. 5e) (Söllner, 2003). Ras-associated binding (RAB) proteins are required in the SNARE complex to tether vesicles to the T-SNARE and allow fusion and secretion (Takahashi et al., 2012). Using an antibody against the small GTPase found on the surface of vesicles, RAB11, we found that it was also localized to the sub-apical region of EOR cells (Fig. 5f), which is the typical docking position for vesicles prior to exocytosis (Söllner, 2003). Taken together, these data suggest that the EOR actively secretes proteins into the lumen of the structure possibly encapsulated in extracellular vesicles.

Presence of SNARE-complex members suggests a role for secretion.

(a) Optical section from confocal Z-stacks of PAX8-rtTa; Tre-H2B-GFP (cyan) EOR at E18.5 using HCR for detection of Clu expression (magenta). (b) Optical section from confocal Z-stacks of PAX8-rtTa; Tre-H2B-GFP (cyan) EOR at E18.5 using HCR for Stxbp2 (magenta). (c) Optical section from confocal Z-stacks of PAX8-rtTa; Tre-H2B-GFP (cyan) EOR at P7/8 using HCR for Clu (magenta). (d) Optical section from confocal Z-stacks of PAX8-rtTa; Tre-H2B-GFP (cyan) EOR at P7/8 using HCR for Stxbp2 (magenta). (e) Optical section from confocal Z-stacks of PAX8-rtTa; Tre-Cre; Rosa-mTmG (cyan) EOR at E18.5 immunostained for STX3 (magenta). Outlined higher resolution image acquired with Airyscan (f) Optical section from confocal Z-stacks of PAX8-rtTa; Tre-Cre; Rosa-mTmG (cyan) EOR at E18.5 immunostained for RAB11 (magenta) Outlined higher resolution image acquired with Airyscan Scale bar – 100um

Proteins produced by the EOR suggest a role in ovary homeostasis

Among proteins captured in our mass spectrometry analysis of the EOR, insulin-like growth factor binding protein 2 (IGFBP2) stood out as a secreted protein with a potential functional role in ovary homeostasis. IGFBP2, found to be secreted by granulosa cells, binds and sequesters IGF1 (Rosenzweig, 2004). The binding of IGFBP2 to IGF1 titrates IGF1 from its receptor IGF1R (Amutha and Rajkumar, 2017). IGF1 has a reported role in ovarian function by amplifying the hormonal action of gonadotropins to promote steroidogenesis and granulosa cell proliferation (Talia et al., 2021).

Because no validated antibodies were commercially available to visualize protein expression for IGFBP2, we again used HCR to determine when and where Igfbp2 was expressed in the epithelial cells of the RO. We found that Igfbp2 was highly expressed in the EOR at E18.5 (Fig. 6a). The mRNA was still present, but at lower levels in P7 EOR cells (Fig. 6b).

Validation of Igfbp2 expression in the RO.

(a) Maximum intensity projection of confocal Z-stacks of PAX8-rtTa; Tre-H2B-GFP (cyan) EOR at E18.5 using HCR for detection of IGFBP2 expression (magenta). (b) Maximum intensity projection of confocal Z-stacks of PAX8-rtTa; Tre-H2B-GFP (cyan) EOR at P7/8 using HCR for detection of IGFBP2 expression (magenta). Bottom panels are outlined higher resolution image acquired with Airyscan. Scale bar top panel – 100um. Scale bar top panel – 25um

Discussion

The female reproductive tract is often thought to consist of the vulva, vagina, cervix, uterus, oviduct, and ovaries (Rendi et al., 2012). We suggest that the RO be added to this list and investigated as an additional component of female reproductive function. We show that the RO consists of 3 distinct regions that mature during fetal life and persist into adulthood. The cells of the EOR show secretory characteristics consistent with the presence of secreted proteins in the lumen identified through mass spectrometry. Importantly, we show that contents of the EOR are transported to the ovary by P7, suggesting a role in ovary homeostasis.

Historically, the regions of the RO were identified using histological sections, where the EOR was defined as a columnar epithelium, while the CR was defined as pseudo-columnar. Recently we found that the entire RO expresses PAX8+, which is usually considered a marker of urogenital epithelial identity. We also showed that the IOR contained PAX8+/FOXL2+ cells and that the EOR was KRT8+ at E17.5 (McKey et al., 2022). In this study we further characterized KRT8 and E-Cadherin expression throughout RO development as the CR acquires epithelial characteristics. We also identified a marker, GFRa1, that is exclusively expressed in the CR. These molecular distinctions predict that the different regions of the RO are functionally distinct.

While PAS-positive staining in the lumen of the EOR of sheep indicated the presence of glycogens, glycoproteins, and proteoglycans (Stein and Anderson, 1979), it remained unclear whether these were secreted by the EOR cells or were accumulated secretions from the ovary. Using injections of a membrane impermeable dye, Dextran, into either the RO or ovary, we found that the fluid within the EOR travels towards the ovary. We also found that the EOR is ensheathed within a layer of aSMA+ mesenchyme that expresses the contractile component Calponin (CNN1) (Gao et al., 2014). Our data suggests that the EOR gains contractility during development, which could facilitate fluid movement within the EOR lumen in conjunction with cilia within the EOR. We found that the ciliary protein ARL13b was expressed in the EOR. While the primary cilium is not motile and does not generate flow, we found in our ScRNA-seq data that cells in the EOR express Pkd2. In renal epithelial cells, nonmotile primary cilia expressing PDK2 sense shear stress during fluid flow and transduce this sensory information (Nauli et al., 2003). Perhaps fluid movement within the EOR is a two-part process, where the primary cilia sense pressure within the lumen inducing contraction of the smooth muscle surrounding the EOR to promote fluid movement towards the ovary. Our E16.5 ScRNA-seq data indicates that the EOR only expresses Arl13b and does not express the multicilia marker Foxj1. However, both markers are present in the putative EOR cell clusters in our adult SnRNA-seq data (Anbarci et al., 2023) suggesting that, after maturation, the EOR likely includes Arl13b+/Foxj1+ multiciliated cells. It is well documented that adult oviductal multiciliated cells express both ARL13b and FOXJ1 (Coy et al., 2016). It is possible that the primary cilia give rise to motile multiciliated cells, as shown in the airway epithelium (Jain et al., 2010) and suggested in the oviduct (Shi et al., 2014).

We used a “milking” process to extrude the contents of EOR, and mass spectrometry, to identify 4,232 proteins. We expected this group to be a mix of proteins from the luminal fluid, and cytoplasmic proteins from lysed EOR cells or from cells closely associated with the EOR. We took a very conservative approach to identify proteins highly likely to be secreted from EOR cells: (1) we compared the proteomics data with the known mammalian secretome to identify only secreted proteins; and (2) we next compared this shortened list with our available E16.5 ScRNA-seq data to find proteins whose transcripts were specific to the EOR at that stage. Although we recognize that the secretome at P7 is likely far more complex, this candidate list of 15 proteins reveals the first insight into specific proteins secreted from the RO that may be involved in the secretory process itself or may be transported to the ovary to affect ovary function.

Within the protein candidate list, we focused on validating clusterin (CLU) and syntaxin binding protein 2 (STXBP2) due to their roles in protein and vesicle transport. Consistent with our proteomic results, both Clu and Stxbp2 RNAs were detected in the RO. CLU is a chaperone protein that is present in both reproductive and non-reproductive tissues (Argraves and Morales, 2004; Saewu et al., 2017). In the epididymis, CLU acts as a chaperone to direct luminal proteins to the sperm head surface (Saewu et al., 2017). Although the role of CLU in the RO remains unknown, we hypothesize that it is important for post-secretory protein guidance to the ovary. STXBP2 binds t-SNARE protein STX and is essential for the vesicle-apical surface membrane fusion stage of vesicle secretion (Söllner, 2003). The presence of STXBP2 in the candidate list, coupled with validation of the expression of Stxbp2 RNA in the cells of the EOR, led us to investigate the presence of other components of the SNARE complex in the EOR.

The SNARE complex is comprised of a vesicle bound v-SNARE (Vamps), membrane bound t-SNARE (STXs), small GTPases (RABs), and syntaxin binding proteins (STXBPs) (Söllner, 2003). Using IF staining, we confirmed the presence of STX3, as well as RAB11. Our ScRNA-seq data indicate that EOR cells also express Vamp7 and Vamp8, which are v-SNAREs typically associated with STX3 (Dingjan et al., 2018). Small GTPases, such as RAB11, are essential in the SNARE complex as they mediate the tethering of vesicle membranes to the apical membrane prior to membrane fusion. Taking into consideration the presence of all components of the SNARE complex, we hypothesize that the EOR utilizes the SNARE complex to promote apical secretion into the luminal fluid.

A protein that was prominent on the proteomic candidate list was IGFBP2. We found that Igfbp2 was highly expressed in both the embryonic and adult RO transcriptome (Anbarci et al., 2023). IGFBP2 is an ideal candidate protein to mediate ovarian function. The IGF1 pathway is essential for regulation of follicular growth and selection (Talia et al., 2021). IGFBP2 can bind and sequester IGF1 (Rosenzweig, 2004), limiting its availability to promote follicle growth through activation of its receptor IGF1R (Amutha and Rajkumar, 2017; Bezerra et al., 2018; Liu et al., 2021). We spatially validated the expression of Igfbp2 in the EOR using HCR, and noted high expression at E18.5, which was decreased by P7. This decrease correlated with the end of the first-wave of follicle activation immediately after birth (Mork et al., 2012; Zheng et al., 2014). We also found variation in levels of Igfbp2 in our adult SnRNA-seq data, where Igfpb2 was enriched specifically in RO cells during the estrus stage compared to the diestrus stage of the estrous cycle. Based on these results, we hypothesize that high levels of estradiol during estrus trigger the EOR to produce IGFBP2 to sequester free IGF1 and decrease follicle growth and production of estradiol, maintaining ovary homeostasis.

The EOR is highly integrated with the extra-ovarian environment. We found that EOR cells are tightly surrounded in a dense web of vasculature. The RO may send or receive information through the vasculature, similar to other epithelial tissues in the female reproductive tract (Reynolds et al., 2002), raising the possibility that the RO participates in endocrine signaling. We also found that the EOR is directly contacted by TUJ1+ neurons, another avenue through which information may be entering or exiting the EOR. It is unclear whether the innervation contacts the epithelial cells specifically, similar to the way in which innervation contacts epithelial cells in the gut (Bohórquez et al., 2015), or whether the innervation contacts the smooth muscle cells that surround the EOR to induce contraction, similar to the innervation-muscle interaction in the uterus (Morizaki et al., 1989). We also found that the EOR is specifically associated with F4/80+ LYVE1-macrophages. LYVE1 macrophages are predicted to have an angiogenic role in the adipose tissue surrounding the epididymis and could possibly be playing a similar role in the EOR (Cho et al., 2007). The integration of the EOR with the extra-ovarian environment, including vasculature, neurons, and macrophages, places the EOR in an ideal position to act as an antenna to interpret homeostatic signals and send information to the ovary. In Drosophila and C. Elegans, there is ample evidence that the ovary responds to physiologic levels of glycogen, insulin, amino acid levels, and changes in diet (Hubbard et al., 2013; Laws and Drummond-Barbosa, 2017). The mediator of this response has not been determined in mammals, but we hypothesize that this is the function of the RO. Ongoing work will investigate how the EOR responds to hormones and other physiological signals, and whether proteins secreted by the EOR such as IGFBP2 respond to physiologic stimuli such as diet or immune status, and convey this information to the ovary.

Methods and Materials

Key Resource Table

Mice and Tissue Collection

Unless otherwise stated, mice used for experiments were maintained on the CD-1 or mixed CD-1 and C57BL/6J genetic backgrounds. The Pax8-rtTA and Tre-H2BGFP lines were previously described (Traykova-Brauch et al., 2008; Tumbar et al., 2004) and maintained on a mixed CD-1/C57BL/6J background, and maintained on a mixed CD-1/C57BL/6J background. The Pax8-rtTa;Tre-Cre;mTmG line was obtained by crossing the Pax8-rtTa line with carriers of the Tre-Cre (Perl et al., 2002) and mTmG (Muzumdar et al., 2007) alleles, which allowed for visualization of cell membranes of PAX8+ cells. Pregnant and nursing dams were given a doxycycline diet at 625 mg/kg (Teklad Envigo TD.01306) 3 days prior to tissue collection to induce GFP expression in Pax8+ cells. Toe samples were collected from mice for genotyping. Primers used for PCR genotyping are listed in the supplementary table 1. To obtain samples at specific stages of development, males were housed with females for timed matings. Successful mating was determined by the presence of a vaginal plug. Date of plug was considered embryonic day 0.5. Tissue samples were collected in phosphate-buffered saline without calcium or magnesium (PBS -/-), fixed in 4% paraformaldehyde (PFA)/PBS for 30 minutes at room temperature and dehydrated stepwise into 100% methanol, followed by storage at -20°C. All mice were housed in accordance with National Institutes of Health guidelines, and experiments were conducted with the approval of the Duke University Medical Center Institutional Animal Care and Use Committee (protocol #: A078-23-03).

Immunostaining and Confocal Image Acquisition

Samples were stepwise rehydrated into 100% PBS, followed by a 30-minute permeabilization wash in PBS 0.1% Triton X-100. Samples were then blocked for 1 hour with PBS, 1% Triton X-100, 10% horse serum, and 3% BSA. Samples were incubated overnight in primary antibodies diluted in blocking solution at 4°C (key resource table). Samples then underwent three 30 - minute washes in permeabilization solution and incubated overnight in secondary antibodies (1:500 dilution) and Hoechst vital dye solution diluted in blocking solution. The next day, samples underwent three 20-minute washes in permeabilization solution before being transferred into 100% PBS and stored at 4°C until ready for confocal imaging. Samples were mounted in DABCO mounting solution and stored at -20°C until imaged. Samples were imaged from both the dorsal and ventral sides using 3D-printed reversible slides that utilize two coverslips that allow for flipping. The 3D model can be downloaded on the NIH 3D Print Exchange website at https://3dprint.nih.gov/discover/3DPX-009765. Samples were imaged in toto using laser scanning confocal microscopy captures in the longitudinal plane on Zeiss LSM780 or LSM880 and affiliated Zen software Carl Zeiss, Inc, Germany) using 10x, 20x, and 63x (also used for Airyscan) objectives.

Hybridization chain reaction

The mouse embryo protocol from Molecular Instruments whole-mount (https://files.molecularinstruments.com/MI-Protocol-RNAFISH-Mouse-Rev9.pdf) was adjusted for P0 and P7 ovary samples. Samples were stepwise rehydrated into 0.1% Tween 20 / PBS (PBST). Samples were then subjected to 10 μg/mL proteinase K solution for 10 minutes at room temperature, then washed twice in PBST for 5 minutes. Samples were then post-fixed in 4% PFA for 10 minutes at room temperature, followed by three PBST washes for 5 minutes. Samples were then pre-hybridized for 30 minutes in 500ul of hybridization buffer (Molecular Instruments) at 37°C. Samples were then incubated overnight at 37°C with HCR probes diluted in 500ul hybridization buffer (2pmol in 500ul). Samples were then washed for 15 minutes four times in the Molecular Instruments probe wash buffer at 37°C, followed by two 5x 0.1% Tween 20/SSC (SSCT) washes for 5 minutes at room temperature. Fluorescently labeled hairpins were snap cooled and left in a dark drawer at room temperature for 30 minutes. Following hairpin preparation, samples were incubated with hairpins diluted in Molecular Instruments amplification buffer overnight in the dark at room temperature. On the third day, samples were washed in SSCT four times for 15 minutes each before being transferred to PBS -/- and then mounted in DABCO mounting medium for imaging.

Image Processing

Confocal images were imported into FIJI software for minor image processing (cropping, rotations, maximum intensity projection/ optical slice montage, and color application). The RO was oriented to the left of the ovary with the oviduct at the top of the ovary. Images were then imported into Adobe Photoshop CC (Adobe, Inc, CA) for final processing of channel overlay, brightness, contrast adjustment, and modification of red to “magenta” (hue adjustment to -30). Channel color accessibility was determined via Photoshop colorblind proofing.

Dextran Injections

Postnatal day 7 mice were euthanized, and the entire ovarian complex was dissected in PBS +/+. Samples were then placed on agar blocks (1.5%) soaked in PBS +/+. The distal tip of the EOR was identified and punctured using a tungsten needle (0.001mm tip diameter, Fine Science Tools, 10130 -05). A microinjection unit (Picospritzer III microinjector) and fine capillary glass needle (5-10 um) loaded with 2-3 nl dextran solution (1.25mg/ml dextran, PBS, 0.05% phenol red) was used to inject dextran into the opening of the EOR. Samples were then left for 15 minutes before fixing with 4% PFA. During the 15 minutes, samples were monitored to visualize movement. Samples were then fixed, stained, and imaged.

Luminal Fluid Collection for Proteomics and Analysis

Postnatal day 7 EORs were collected from Pax8-rtTa; Tre-H2B-Gfp mice (as a guide to only collect EOR) and carefully cleaned up to remove as much non-EOR tissue (40 EOR samples). Samples were then placed in a 1.5ml Eppendorf tube and ‘pressed’ for fluid using a disposable pellet pestle. The sample was then spun down and the supernatant was collected and snap frozen in liquid nitrogen. The sample was then submitted to the Duke University Proteomics Core for mass spectrometry. Proteomic analysis uncovered 4,252 proteins present in the pressed fluid. To exclude proteins that may have contributed due to cell lysis, results were cross-referenced against a database of secreted proteins (Meinken et al., 2015). Overlapping proteins were then cross-referenced to our E16.5 ScRNA-seq data (Anbarci et al., 2023), and candidate proteins were determined by gene expression specific to the EOR cluster.

LC-MS/MS Proteomics Analysis

The sample was subjected to a Bradford (Pierce) protein measurement and 10 ug was removed for downstream processing. The sample was brought to 4% SDS, reduced with 10 mM dithiothreitol for 20 min at 55°C, alkylated with 25mM iodoacetamide for 45 min at room temperature and then subjected to S-trap (Protifi) trypsin digestion using manufacturer recommended protocols. Digested peptides were lyophilized to dryness and resuspended in 50 uL of 0.2% formic acid/2% acetonitrile. The sample was subjected to chromatographic separation on a Waters MClass UPLC equipped with a 1.8 μm Acquity HSS T3 C18 75 μm × 250 mm column (Waters Corp.) with a 90-min linear gradient of 5 to 30% acetonitrile with 0.1% formic acid at a flow rate of 400 nanoliters/minute (nL/min) with a column temperature of 55°C. Data collection on the Fusion Lumos mass spectrometer with a FAIMS Pro device was performed for three difference compensation voltages (-40v, -60v, -80v). Within each CV, a data-dependent acquisition (DDA) mode of acquisition with a r=120,000 (@ m/z 200) full MS scan from m/z 375 – 1500 with a target AGC value of 4e5 ions was performed. MS/MS scans with HCD settings of 30% were acquired in the linear ion trap in “rapid” mode with a target AGC value of 1e4 and max fill time of 35 ms. The total cycle time for each CV was 0.66s, with total cycle times of 2 sec between like full MS scans. A 20s dynamic exclusion was employed to increase depth of coverage. The total analysis cycle time for each sample injection was approximately 2 hours.

Raw LC-MS/MS data files were processed in Proteome Discoverer 3.0 (Thermo Scientific) and then submitted to independent Sequest database searches against a Mus musculus protein database containing both forward and reverse entries of each protein. Search tolerances were 2 ppm for precursor ions and 0.8 Da for product ions using trypsin specificity with up to two missed cleavages. Carbamidomethylation (+57.0214 Da on C) was set as a fixed modification, whereas oxidation (+15.9949 Da on M) was considered a dynamic mass modifications. All searched spectra were imported into Scaffold (v5.3, Proteome Software) and scoring thresholds were set to achieve a peptide false discovery rate of 1% using the PeptideProphet algorithm. Data were output as total spectral matches.

Supporting information

eLife Assessment

This important study reports the developmental dynamics and molecular markers of the rete ovarii during ovarian development. However, the data supporting the main conclusions remain incomplete. This study will be of interest to developmental and reproductive biologists.

Reviewer #1 (Public Review):

Summary:

The manuscript by Anbarcia et al. re-evaluates the function of the enigmatic Rete Ovarii (RO), a structure that forms in close association with the mammalian ovary. The RO has generally been considered a functionless structure in the adult ovary. This manuscript follows up on a previous study from the lab that analyzed ovarian morphogenesis using high-resolution microscopy (McKey et al., 2022). The present study adds finer details to RO development and possible function by (1) identifying new markers for OR sub-regions (e.g. GFR1a labels the connecting rete) suggesting that the sub-regions are functionally distinct, (2) showing that the OR sub-regions are connected by a luminal system that allows transport of material from the extra-ovarian rete (EOR) to the inter-ovarian rete (IOG), (3) identifies proteins that are secreted into the OR lumen and that may regulate ovarian homeostasis, and finally, (4) better defines how the vasculature, nervous, and immune system integrates with the OR.

Strengths:

The data is beautifully presented and convincing. They show that the RO is composed of three distinct domains that have unique gene expression signatures and thus likely are functionally distinct.

Weaknesses:

It is not always clear what the novel findings are that this manuscript is presenting. It appears to be largely similar to the analysis done by McKey et al. (2022) but with more time points and molecular markers. The novelty of the present study's findings needs to be better articulated.

Reviewer #2 (Public Review):

A large number of ovarian experiments have been conducted - especially in morphological and molecular biology studies - specifically removing the ovarian membrane. This experiment is a good supplement to existing knowledge and plays an important role in early ovarian development and the regulation of ovarian homeostasis during the estrous cycle. There are also innovations in research ideas and methods, which will meet the requirements of experimental design and provide inspiration for other researchers.

This reviewer did not identify any major issues with the article. However, the following points could be further clarified:

(1) Is there any comparative data on the proteomics of RO and rete testis in early development? With some molecular markers also derived from rete testis, it would be better to provide the data or references.

(2) Although the size of RO and its components is quite small and difficult to operate, the researchers in this article had already been able to perform intracavitary injection of EOR and extract EOR or CR for mass spectrometry analysis. Therefore, can EOR, CR, or IOR be damaged or removed, providing further strong evidence of ovarian development function?

(3) Although IOR is shown on the schematic diagram, it cannot be observed in the immunohistochemistry pictures in Figure 1 and Figure 3. The authors should provide a detailed explanation.

Reviewer #3 (Public Review):

Summary:

The rete ovarii (RO) has long been disregarded as a non-functional structure within the ovary. In their study, Anbarci and colleagues have delineated the markers and developmental dynamics of three distinct regions of the RO - the intraovarian rete (IOR), the extraovarian rete (EOR), and the connecting rete (CR). Notably focusing on the EOR, the authors presented evidence illustrating that the EOR forms a convoluted tubular structure culminating in a dilated tip. Intriguingly, microinjections into this tip revealed luminal flow towards the ovary containing potentially secreted functional proteins. Additionally, the EOR cells exhibit associations with vasculature, macrophages, and neuronal projections, proposing the notion that the RO may play a functional role in ovarian development during critical ovariogenesis stages. By identifying marker genes within the RO, the authors have also suggested that the RO could serve as a potential structure linking the ovary with the neuronal system.

Strengths:

Overall, the reviewer commends the authors for their systematic research on the RO, shedding light on this overlooked structure in developing ovaries. Furthermore, the authors have proposed a series of hypotheses that are both captivating and scientifically significant, with the potential to reshape our understanding of ovarian development through future investigations.

Weaknesses:

There is a lack of conclusive data supporting many conclusions in the manuscript. Therefore, the paper's overall conclusions should be moderated until functional validations are conducted.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Weaknesses:

It is not always clear what the novel findings are that this manuscript is presenting. It appears to be largely similar to the analysis done by McKey et al. (2022) but with more time points and molecular markers. The novelty of the present study's findings needs to be better articulated.

The previous study focused on placing the Rete Ovarii in the context of ovarian development. The current study focuses on the novel findings that the EOR is a active structure that sends fluid/information to the ovary. We show this by characterizing the presence of secretory proteins in the RO epithelial cells, by dye injections into the EOR and observing transport of the dye to the ovary, and by collection of EOR fluid followed by proteomic analysis. We also show that RO is embedded in an elaborate vascular network and contacted by neurons. None of this data was not discussed in the McKey 2022 paper.

Reviewer #2 (Public Review):

Clarifications:

(1) Is there any comparative data on the proteomics of RO and rete testis in early development? With some molecular markers also derived from rete testis, it would be better to provide the data or references.

To the best of our knowledge, there are no available proteomic datasets of the embryonic or early postnatal mouse Rete Testis or Epididymis. The authors agree that having this information would be very useful.

(2) Although the size of RO and its components is quite small and difficult to operate, the researchers in this article had already been able to perform intracavitary injection of EOR and extract EOR or CR for mass spectrometry analysis. Therefore, can EOR, CR, or IOR be damaged or removed, providing further strong evidence of ovarian development function?

We attempted to genetically ablate the RO by expressing the diphtheria toxin receptor (DTR) in RO cells and adding DT. This approach was not successful in ablating the RO. We also tried to use Pax2/8 homo- and heterozygous mutants for ablation (as used in the McKey 2022 paper), but so far, we cannot find a genetic combination that ablates the RO, but not the oviduct, uterus and/or kidneys. We have also embarked on a study to surgically remove the RO. This assay is taking some time to optimize. The goal of the current study was to characterize the cells along the length of the RO and to present evidence that it is a secretory appendage of the ovary.

(3) Although IOR is shown on the schematic diagram, it cannot be observed in the immunohistochemistry pictures in Figure 1 and Figure 3. The authors should provide a detailed explanation.

An annotation has been added to Figure 1 to indicate the IOR. As the images within the panels are of maximum intensity projections, it is often difficult to clearly see the IOR as it is deeper within the ovary. In Figure 3, the view of the ovary is from the ventral side: this view does not allow for clear visualization of the IOR.

Reviewer #3 (Public Review):

Weaknesses:

There is a lack of conclusive data supporting many conclusions in the manuscript. Therefore, the paper's overall conclusions should be moderated until functional validations are conducted.

We have moderated the conclusions where appropriate

Reviewer #1 (Recommendations For The Authors):

(1) The introduction is relatively brief and does not mention some historical data/hypotheses on the role of the RO in ovarian function (e.g. regulation of meiotic entry) or development (e.g. Mayère et al., 2022).

Mayere 2022 was cited in line 57. Steins hypothesis about entry into meiosis has been added line 58.

(2) L82-84: It is stated that KRT8 was first identified as a potential RO marker by sc/snRNAseq (Anbarci et al., 2023) and then validated in this manuscript. However, KRT8 was used by McKey et al. (2022) as a RO marker, and they noted there that KRT8 was enriched in the EOR. It is not clear why McKey et al. is not cited as the primary reference validating KRT8 as an EOR marker.

The embryonic and neonatal timecourse description from KRT8 expression is first identified in this paper. McKey 2022 only highlights KRT8 at E18.5 A reference has been added to address this line 85

(3) Figure 1: Can the IOR be seen in these images? If so, please label.

The label has been added.

(4) L107: It is hypothesized that "the RO may respond to or interpret homeostatic cues." Can transcriptomics data shed light on what signals the RO may be capable of responding to? E.g. what receptors are expressed by cells of the RO (e.g. ER, LHCGR, FSHR)?

The RO expresses ESR1, PGR, INSR, IGF1R. The IOR exclusively expresses LHCGR and FSHR.This has been added to the manuscript line 309

(5) L152: Mass spec was used to identify proteins secreted into the lumen of the RO. These proteins were then compared to the mammalian secretome to filter out possible nonsecreted protein contaminants. Finally, the candidates were compared to the RO scRNAseq data from Anbarci et al., (2023). This method gives a very conservative candidate list. However, it may also be informative to compare the sc/snRNA-seq gene list directly to the secretome to ID other possible candidate-secreted proteins that may not have been detected in the mass spec data set.

There are quite a number of secreted proteins that are also not actively secreted. This is a good suggestion for future analysis. For the current study we wanted to take a more conservative approach, and chose to do proteomics to determine proteins that are actively secreted.

(6) L195: It is not clear if IGFBP2 is expressed by both OR and granulosa cells or only granulosa cells. It would be informative to know what ovarian cell types express both IGFBP2 and IGF1R (e.g. from sc/snRNA-seq)? This information is referenced in the discussion (L285-287) but would be better to reference it in the results section for clarity.

Both RO and granulosa cells express IGFBP2 and IGF1R. A sentence has been added to results for clarity. (Line 197)

(7) L295: "...the RO participates in endocrine signaling..." might be more accurate to say "...the RO responds to endocrine signaling...".

The authors agreed that this statement is more accurate and the changes have been made.

Reviewer #3 (Recommendations For The Authors):

Several issues significantly affect the paper's quality in the current version. Firstly, there is a lack of conclusive data supporting many conclusions in the manuscript. For instance, the assertion in line 105 that "EOR was directly innervated by neurons" lacks substantial evidence beyond basic immunofluorescent staining.

We agree that the term “innervated” might be a step too far since we rely on IF evidence. We changed the wording of this sentence to say, “The EOR was directly contacted by neurons”.

In another pivotal experiment illustrated in Figure 3, the provided images lack temporal continuity and quantitative analysis, suggesting the incorporation of time-lapse imaging for improved sequential presentation in Figure 3.

The microscope where we can perform injections cannot record movies. We have tried moving the rete to another microscope after injection, but so far, we have been unable to capture dextran moving through the RO. We therefore believe that transport is rapid, but future experiments will be needed to optimize this imaging.

Moreover, relying solely on proteomics analysis, as seen in lines 188-189, makes it challenging to assert conclusions such as "EOR actively secretes proteins." Therefore, the paper's overall conclusions should be moderated until functional validations are conducted.

The findings that (1) the cells of the EOR express SNARE complex proteins at their apical surfaces and (2) luminal fluid expelled from the EOR contains abundant secreted proteins strongly suggest that the RO is involved in active secretion. We use the word “suggest” in this sentence, lines 188-189 as we realize that further experiments should be done to validate this conclusion.

Furthermore, the predominant methods in this study involve immunostaining and imaging. However, the current images exhibit a notable inconsistency in color definitions for different markers by the authors. For instance, in Figure 2.A/C, PAX8 is portrayed as cyan, while in D, it is represented in yellow. Similarly, in Figure 4, E-CAD is depicted using both cyan and yellow. Utilizing different colors for the same protein within a figure can significantly confuse readers' interpretation of the experiments. Rectifying these inconsistencies is essential to enhance the clarity and comprehension of the experimental results.

These colors were chosen to be visible to those with color image impairments. We typically used cyan and magenta to emphasize the most important markers in the image. When E-Cad and KRT8 were often used to emphasized or landmark a structure by localization of these protein. When KRT8 and E-Cad were highlighted, they were represented in cyan and magenta for visibility. When these proteins were used as a landmark to orient the reader and not as the main point, they were labeled in yellow.

At last, many markers in this study are derived from bulk and single-cell sequencing of developing RO. However, it seems that these important data were separated into another paper as a preprint. If this data were incorporated into the current manuscript, the manuscript would become more comprehensive for guiding future research on the RO.

Since we have single cell and single nuclei data from fetal and adult estrus and metestrus stages, we found that incorporating all this data into the present manuscript was overwhelming. Instead, we devoted another manuscript to presenting and validating that data. We believe a quick look at the sequencing manuscript will make this clear.

Version 2

Abstract

Abstract

The rete ovarii (RO) is an appendage of the ovary that has been given little attention. Although the RO appears in drawings of the ovary in early versions of Gray’s Anatomy, it disappeared from recent textbooks, and is often dismissed as a functionless vestige in the adult ovary. Using PAX8 immunostaining and confocal microscopy, we characterized the fetal development of the RO in the context of the ovary. The RO consists of three distinct regions that persist in adult life, the intraovarian rete (IOR), the extraovarian rete (EOR), and the connecting rete (CR). While the cells of the IOR appear to form solid cords within the ovary, the EOR rapidly develops into a convoluted tubular epithelium ending in a distal dilated tip. Cells of the EOR are ciliated and exhibit cellular trafficking capabilities. The CR, connecting the EOR to the IOR, gradually acquires tubular epithelial characteristics by birth. Using microinjections into the distal dilated tip of the EOR, we found that luminal contents flow towards the ovary. Mass spectrometry revealed that the EOR lumen contains secreted proteins potentially important for ovarian function. We show that the cells of the EOR are closely associated with vasculature and macrophages, and are contacted by neuronal projections, consistent with a role as a sensory appendage of the ovary. The direct proximity of the RO to the ovary and its integration with the extraovarian landscape suggest that it plays an important role in ovary development and homeostasis.

Introduction

The Rete Ovarii (RO) is an epithelial structure that is directly connected to the ovary, first described over a century ago (Waldeyer, 1870) as a multi-region structure of mesonephric origin. Despite high conservation among mammalian species including guinea pigs, cows, cats, sheep, swine, prairie deer mice, camels, dogs, monkeys (Cassali et al., 2000; Kim et al., 2012), cows (Santos et al., 2012), and humans (Khan et al., 1999), previous work on the RO did not arrive at a consensus on the function of this structure. Thus, it has remained a mysterious orphan structure.

Although the RO has been proposed to be the female homologue of the rete testis (Wenzel and Odend’hal, 1985), our recent work revealed that it is more complex (McKey et al., 2022). The RO is composed of three distinct regions: (1) the extraovarian rete (EOR) which consists of columnar epithelial cells that create a single convoluted tubule structure ending in a blind distal dilated tip (DDT), (2) the connecting rete (CR) which consists of pseudo-columnar cells; and (3) the intraovarian rete (IOR) which consists of squamous epithelial cells that branch and form a fine network of thin solid cell cords approximately 1-3 cells thick (Byskov and Lintern-Moore, 1973; McKey et al., 2022). It was previously shown that the RO consists of ciliated and non-ciliated cells rich in apical microvilli and mitochondria (Czernobilsky et al., 1985). Fine granular Periodic acid - Schiff (PAS) positive material was found in the cytoplasm of the EOR and CR cells, indicating the presence of polysaccharides (e.g. glycogen and mucins) (Stein and Anderson, 1979). Because the proportion of PAS-positive material in the RO was found to be influenced by the estrous cycle in cows (Wenzel et al., 1987), researchers concluded that the luminal contents of the RO were under endocrine control (Archbald et al., 1971). Although these results strongly suggested a secretory role for the RO, this function was not experimentally confirmed, and no proteomic or metabolomic investigations were pursued. Previous characterization of the RO relied heavily on imaging serial sections to study the structure. However, these studies lacked contextual information, and the use of sections made it challenging to determine whether tubules were connected to one another or were isolated structures (Wenzel and Odend’hal, 1985; Woolnough et al., 2000). Furthermore, some investigators reported that the whole RO contracted postnatally and that the DDT of the EOR separated from the RO and degenerated (Byskov and Lintern-Moore, 1973). Perhaps for these reasons, the RO was deemed a functionless vestige, and has been omitted from recent textbook representations of the female reproductive tract (Girsh, 2021).

The RO was rediscovered and highlighted in our recent study (McKey et al., 2022), where we used confocal and lightsheet imaging of whole ovaries to study the integration of ovary morphogenesis with the development of surrounding tissues, including the RO. Recently, the IOR has gained attention as a newly described progenitor for supporting cells of the murine gonad (Mayère et al., 2022). This echoes previous data suggesting that the RO plays a role in the onset of meiosis (Stein and Anderson, 1979). Recent single cell transcriptomics studies have also identified cells of the RO within human fetal gonads (Lardenois et al., 2023; Taelman et al., 2022). We reported that the entire RO expresses high levels of PAX8, and used this as a marker to visualize and characterize cells of the RO. In the present study, we used the Pax8-rtTa; Tre-H2B-Gfp RO nuclear reporter mouse line, advanced imaging techniques, and secretome analysis to characterize the development of the intact RO in its native context and to investigate the function and heterogeneity of its cells. We found that the RO arises from a subset of the mesonephric tubules, analogous to the rete testis, and persists into adulthood. Our studies reveal that the RO is a continuous structure, surrounded by smooth muscle actin, a dense vascular network, and several macrophage populations. We also show that the RO is directly contacted by neurons. The enrichment of secretory machinery in RO cells as well as our experimental analysis of directional flow and luminal contents, together suggest that the RO sends material to the ovary. Based on these findings, we suggest that the RO plays a role in ovary function and should be investigated as a functional organ of the female reproductive tract.

Results

Characterization of the IOR, CR, and EOR

We used immunofluorescence (IF) and confocal imaging to investigate the development and sub-regional structure of the RO from embryonic day (E) 16.5 to postnatal day (P) 7. Three distinct regions of the RO were originally defined histologically by cell morphology (Byskov and Lintern-Moore, 1973; Lee et al., 2011). Using IF and confocal imaging, we found that these three regions are maintained throughout development, but their relative sizes change. While the entirety of the RO is PAX8+, we took different approaches to identify region specific markers. First, as GFRa1 was known to be expressed in the rete testis, we hypothesized that it was expressed in the RO as well. Indeed, immunostaining using antibodies against GFRa1 revealed that part of the RO is GFRa1+. GFRa1 specifically labeled the CR from E16.5 to P7 (Fig. 1 bottom row). To identify additional RO markers, we performed bulk and single cell RNA sequencing (ScRNA-seq) of cells from the ovarian complex of mice at E16.5 and 2 months and found that Krt8 was enriched in the RO (McKey et al., 2022;Anbarci et al., 2024). We validated these findings using IF against KRT8, which revealed that KRT8+ cells were specifically localized to the EOR (Fig. 1 third row) (McKey et al., 2022).

The Rete Ovarii undergoes dynamic changes during development alongside the ovary.

(a,b,c,d top row) Diagrams depicting the development of the three regions (EOR-magenta; CR-cyan; IOR-yellow) of the RO from E16.5-P7 in the whole ovarian complex (ovary-light pink; oviduct-grey). (a,b,c,d bottom rows) Maximum intensity projection from confocal Z-stacks of whole ovary/mesonephros complexes at E16.5 (a), E18.5 (b), P0 (c) and P7 (d) immunostained for PAX8 (yellow), GFRa1 (cyan) and KRT8 (magenta). a,b,c,d second row are composite images, while the third row show separate panels for KRT8 and and the bottom row show panels for GFRa1. *Note that GFRa1 and KRT8 do not co-localize and are specific to the CR and EOR, respectively. All figures are dorsal views of the ovary. Yellow asterisk indicates opening of the infundibulum for reference. Scale bar – 100um

To characterize RO development, we next analyzed PAX8 and KRT8 expression from E16.5 to P0. At E16.5, the CR and IOR were the largest of the three regions. In the EOR, several tubules leading from the region of the CR converged into a single tube of columnar PAX8+ and KRT8+ epithelial cells that lead to a blind end. Between E16.5 and E18.5, the EOR underwent rapid expansion, and the blind end of the EOR became dilated. This structure, which we refer to as the distal dilated tip (DDT) was best visualized from the ventral side of the ovary (SFig. 1). By E18.5, the EOR was the largest of the three regions. At this stage the CR was still large, but the IOR began to regress. At P0, the EOR remained the largest region, and the IOR had regressed to the medullary region of the ovary (McKey et al., 2022). Their distinct structure and protein expression patterns suggest that each region has a different function.

Integration of the EOR with the extra-ovarian environment

During gonadogenesis, the mesonephros is highly vascularized in both XX and XY embryos (Cool et al., 2011). In contrast, the ovary and surrounding tissue are more highly innervated than the testis (McKey et al., 2019). A surprising finding in our bulk transcriptome analysis of the RO was the presence of a high proportion of immune cells and cells with vascular markers that co-isolated with E16.5 RO cells (Anbarci et al., 2024). To explore the integration of the RO with its environment, we used IF to investigate the expression of endothelial marker Endomucin (Fig. 2a), smooth muscle marker alpha smooth muscle actin (aSMA) (Fig. 2b), pan-neuronal marker TUJ1 (Fig. 2c), and macrophage markers F4/80 and LYVE1 (Fig. 2d). We found that at E18.5 the EOR was tightly surrounded by vasculature and ensheathed within a layer of aSMA+ mesenchyme. The EOR, and more specifically the DDT, was directly contacted by neurons that contact the PAX8+ epithelial cells. We also found that the EOR is specifically associated with F4/80+ macrophages (Fig. 4d). This multifaceted integration with the environment suggests the RO may respond to or interpret homeostatic cues.

The EOR is highly integrated with its extra-ovarian environment suggesting multifaceted communication.

(a) Top panel is a whole ovarian complex maximum intensity projection of the confocal Z-stack at E18.5 immunostained for ENDOMUCIN (magenta) and PAX8 (cyan). Bottom panel is an optical section showing vasculature tightly surrounding the DDT of the EOR (magenta). (b) Top panel is a whole ovarian complex maximum intensity projection of the confocal Z-stack at E18.5 immunostained for aSMA (magenta) and KRT8 (cyan). Bottom panel is an optical section showing the EOR tightly ensheathed by smooth muscle (magenta). (c) Top panel is a whole ovarian complex maximum intensity projection of the confocal Z-stack at E18.5 immunostained for TUJ1 (magenta) and PAX8 (cyan). Bottom panel is an optical section showing direct contacts between the EOR and neuronal projections (magenta). (d) EOR at P0 immunostained for PAX8 (yellow), F4/80 (magenta) and LYVE-1 (cyan). Top and bottom panel are maximum intensity projection of the confocal Z-stack. Bottom panel shows the absence of LYVE1 macrophages (cyan) proximate to the EOR, in contrast to the closely associated F4/80 macrophages (magenta in top image). Arrowheads show regions devoid of LYVE1 macrophages where F4/80 macrophages are present. Scale bar – 100um

Flow of luminal material within the EOR

Because the EOR lumen is fluid filled, and because the EOR and CR cells are rich in proteoglycans (Byskov, 1975) we hypothesized that the EOR produces luminal secretions. To investigate this idea, we first chose to determine the direction of flow at P7, when the DDT of the EOR was fully dilated (Fig. 1d). We first injected fluorescently labeled pH-insensitive Dextran into the DDT of the EOR and found that, within just 15 min, the fluorescent fluid had readily traveled from the DDT into the ovary, where it then diffused widely (Fig. 3). By contrast, when we injected fluorescently labeled dextran into the P7 ovary near the IOR, the dextran remained in the ovary and did not travel to the EOR (Fig. 3). These data indicate that, at least at P7, the fluid inside the lumen of the EOR travels towards the ovary.

Fluid moves from the EOR to the ovary.

(a,b) Schematic of dextran injections (syringe) for each group (a- EOR injections, b- ovary injections) (Oviduct-grey, ovary-light pink, IOR-yellow, CR-cyan, EOR-magenta). (Second row) Maximum intensity projection of a confocal Z-stack of whole ovarian complexes at P7, where Dextran was injected into the EOR. Presence of Dextran in the ovary shows that when dextran is injected into the EOR, it diffuses throughout the ovary (dextran-cyan, E-CADHERIN-magenta). (bottom row) Maximum intensity projection from confocal Z-stacks of whole ovary/mesonephros complexes at P7 where Dextran was injected into the ovary. Absence of Dextran signal in the EOR shows that dextran did not travel into the EOR when injected into the ovary near the IOR. Scale bar – 100um

Potential of the EOR and CR for fluid transfer

Without the obvious indication of a lumen within the CR at P7 and taking into consideration the absence of the epithelial marker KRT8 in those cells, we wondered how fluid could travel through the CR to the ovary. Using an antibody against E-CADHERIN, a marker of cell junctions, we found that it was specifically present in the EOR at E16.5, but absent from the CR (Fig. 4a, second row). However, E-Cadherin expression was gradually gained in the CR and IOR such that, by P7, the entire RO was positive for this epithelial marker (Fig. 4a-d, second row). These results suggest that tubular epithelial connections between the EOR and IOR mature gradually between E16.5-P7.

The EOR and CR acquire the potential for fluid transfer

(a,b,c,d) Cells of the CR acquire E-CADHERIN soon after birth. Maximum intensity projection from confocal Z-stacks of whole ovary/mesonephros complexes at E16.5-P7 immunostained for PAX8 (magenta) and E-CADHERIN (cyan). Arrowhead shows the region of the CR where E-CADHERIN expression is negligible at E16.5 (a), low at E18.5 (b) and fully expressed at P0 (c) and P7 (d). (e) Optical section from confocal Z-stacks of RO complexes at P0 immunostained for E-CADHERIN (yellow), CNN1 (magenta), and aSMA (cyan). Boxed regions show the absence of CNN1 in the aSMA+ sheath around the EOR at P0, and (f) its acquisition by P7, suggesting gain of contractility. (g) The RO expresses cilia marker ARL13b by P0. Maximum intensity projection from confocal Z-stacks of RO complexes at E18.5 immunostained for E-CADHERIN (cyan) and ARL13b (magenta). Outlined boxes show all regions of the EOR have ciliated cells. Scale bar – 100um (h). Airyscan optical section from confocal Z-stacks of RO complexes at E18.5 immunostained for KRT8 (cyan) and ARL13b (magenta). Scale bar – 25um

Next, we investigated mechanisms that could facilitate fluid movement through the RO. Previous reports showed that ciliated and non-ciliated cells are present in the RO (Lee et al., 2011). Using an antibody against a marker for cilia, ARL13b, we found that primary ciliated cells were abundant in the DDT and throughout the tubules of the EOR at E18.5 (Fig. 4g-h). Due to the curvature of the DDT, optical slices often include cilia from neighboring cells. However, an ultrathin (0.6um) optical section suggested that each cell has a single cilium. Since the EOR was covered in a sheath of aSMA+ mesenchyme (Fig. 2b, Fig. 4e-f), we investigated whether and when this layer of smooth muscle became contractile. We used an antibody against the contractile smooth muscle protein Calponin (CNN1) and found that CNN1 was absent around the EOR during fetal development and in neonates. However, by P7, the mesenchymal sheath around the EOR gained expression of CNN1, indicating that it acquired the ability to contract by this stage of development (Fig. 4e-f). These data suggest that ciliary mechanosensing and/or muscle contraction may aid the directional movement of the fluid from the DDT to the ovary.

Proteins produced by the EOR indicate a role for the SNARE-complex

Our data showing fluid flow from the EOR to the ovary prompted us to investigate the nature and identity of the proteins produced and potentially secreted by the EOR. To address this question, we analyzed the protein contents of the luminal material using mass spectrometry. Using the Pax8rtTA; Tre-H2B-GFP reporter mouse line, we dissected EORs and isolated luminal fluid by gently pressing the tissue with a pestle. Using this method, we anticipated several problems. First, we expected that some cells within the EOR and surrounding tissue would be lysed during this procedure, which would release proteins not ordinarily secreted. Second, we expected that some cells closely associated with the EOR would be co-isolated, which could result in contamination with contents from cells that are not part of the EOR. To eliminate intracellular proteins arising from lysed cells, we cross-referenced our proteomic dataset with that of the mammalian secretome (Meinken et al., 2015), thereby retaining only known secreted proteins. Next, we compared the resulting list to the E16.5 ScRNA-seq data from cells mapping to the RO (Anbarci et al., 2024). This produced a conservative candidate list of 15 proteins (Table 1). Of the candidate proteins, two were selected for validation, CLU and STXBP2, due to their roles in protein and vesicle transport (Argraves and Morales, 2004; Saewu et al., 2017; Söllner, 2003).

Because no validated antibodies were commercially available to visualize protein expression for CLU and STXBP2, validations were performed using Hybridization Chain Reaction (HCR), a method for single - molecule RNA-Fluorescence in situ hybridization (FISH) (https://files.molecularinstruments.com/MI-Protocol-RNAFISH-Mouse-Rev9.pdf). We found that both Clu and Stxbp2 were expressed in the EOR at E18.5 (Fig. 5a-b) and at P7 (Fig. 5c-d). The presence of Stxbp2 showed that components of the SNARE-complex were actively transcribed in the EOR, suggesting this may serve as a mechanism for secretion. The presence of other components of the SNARE-complex were validated using IF. We found that the T-SNARE complex member STX3was expressed throughout the EOR and was localized to the apical surface, a cellular position that is typically associated with active secretion (Fig. 5e) (Söllner, 2003). Ras-associated binding (RAB) proteins are required in the SNARE complex to tether vesicles to the T-SNARE and allow fusion and secretion (Takahashi et al., 2012). Using an antibody against the small GTPase found on the surface of vesicles, RAB11, we found that it was also localized to the sub-apical region of EOR cells (Fig. 5f), which is the typical docking position for vesicles prior to exocytosis (Söllner, 2003). Taken together, these data suggest that the EOR actively secretes proteins into the lumen of the structure possibly encapsulated in extracellular vesicles.

Presence of SNARE-complex members suggests a role for secretion.

(a) Optical section from confocal Z-stacks of PAX8-rtTa; Tre-H2B-GFP (cyan) EOR at E18.5 using HCR for detection of Clu expression (magenta). (b) Optical section from confocal Z-stacks of PAX8-rtTa; Tre-H2B-GFP (cyan) EOR at E18.5 using HCR for Stxbp2 (magenta). (c) Optical section from confocal Z-stacks of PAX8-rtTa; Tre-H2B-GFP (cyan) EOR at P7/8 using HCR for Clu (magenta). (d) Optical section from confocal Z-stacks of PAX8-rtTa; Tre-H2B-GFP (cyan) EOR at P7/8 using HCR for Stxbp2 (magenta). (e) Optical section from confocal Z-stacks of PAX8-rtTa; Tre-Cre; Rosa-mTmG (cyan) EOR at E18.5 immunostained for STX3 (magenta). Outlined higher resolution image acquired with Airyscan (f) Optical section from confocal Z-stacks of PAX8-rtTa; Tre-Cre; Rosa-mTmG (cyan) EOR at E18.5 immunostained for RAB11 (magenta) Outlined higher resolution image acquired with Airyscan Scale bar – 100um

Proteins produced by the EOR suggest a role in ovary homeostasis

Among proteins captured in our mass spectrometry analysis of the EOR, insulin-like growth factor binding protein 2 (IGFBP2) stood out as a secreted protein with a potential functional role in ovary homeostasis. IGFBP2, found to be secreted by granulosa cells, binds and sequesters IGF1 (Rosenzweig, 2004). The binding of IGFBP2 to IGF1 titrates IGF1 from its receptor IGF1R (Amutha and Rajkumar, 2017). IGF1 has a reported role in ovarian function by amplifying the hormonal action of gonadotropins to promote steroidogenesis and granulosa cell proliferation (Talia et al., 2021).

Because no validated antibodies were commercially available to visualize protein expression for IGFBP2, we again used HCR to determine when and where Igfbp2 was expressed in the epithelial cells of the RO. We found that Igfbp2 was highly expressed in the EOR at E18.5 (Fig. 6a). The mRNA was still present, but at lower levels in P7 EOR cells (Fig. 6b).

Validation of Igfbp2 expression in the RO.

(a) Maximum intensity projection of confocal Z-stacks of PAX8-rtTa; Tre-H2B-GFP (cyan) EOR at E18.5 using HCR for detection of IGFBP2 expression (magenta). (b) Maximum intensity projection of confocal Z-stacks of PAX8-rtTa; Tre-H2B-GFP (cyan) EOR at P7/8 using HCR for detection of IGFBP2 expression (magenta). Bottom panels are outlined higher resolution image acquired with Airyscan. Scale bar top panel – 100um. Scale bar top panel – 25um

Discussion

The female reproductive tract is often thought to consist of the vulva, vagina, cervix, uterus, oviduct, and ovaries (Rendi et al., 2012). We suggest that the RO be added to this list and investigated as an additional component of female reproductive function. We show that the RO consists of 3 distinct regions that mature during fetal life and persist into adulthood. The cells of the EOR show secretory capabilities consistent with the presence of secreted proteins in the lumen identified through mass spectrometry. Importantly, we show that contents of the EOR are transported to the ovary by P7, suggesting a role in ovary homeostasis.

Historically, the regions of the RO were identified using histological sections, where the EOR was defined as a columnar epithelium, while the CR was defined as pseudo-columnar. Recently we found that the entire RO expresses PAX8+, which is usually considered a marker of urogenital epithelial identity. We also showed that the IOR contained PAX8+/FOXL2+ cells and that the EOR was KRT8+ at E17.5 (McKey et al., 2022). In this study we further characterized KRT8 and E-Cadherin expression throughout RO development as the CR acquires epithelial characteristics. We also identified a marker, GFRa1, that is exclusively expressed in the CR. These molecular distinctions predict that the different regions of the RO are functionally distinct.

While PAS-positive staining in the lumen of the EOR of sheep indicated the presence of glycogens, glycoproteins, and proteoglycans (Stein and Anderson, 1979), it remained unclear whether these were secreted by the EOR cells or were accumulated secretions from the ovary. Using injections of a membrane impermeable dye, Dextran, into either the RO or ovary, we found that the fluid within the EOR travels towards the ovary. We also found that the EOR is ensheathed within a layer of aSMA+ mesenchyme that expresses the contractile component Calponin (CNN1) (Gao et al., 2014). Our data suggests that the EOR gains contractility during development, which could facilitate fluid movement within the EOR lumen in conjunction with cilia within the EOR. We found that the ciliary protein ARL13b was expressed in the EOR. While the primary cilium is not motile and does not generate flow, we found in our ScRNA-seq data that cells in the EOR express Pkd2. In renal epithelial cells, nonmotile primary cilia expressing PDK2 sense shear stress during fluid flow and transduce this sensory information (Nauli et al., 2003). Perhaps fluid movement within the EOR is a two-part process, where the primary cilia sense pressure within the lumen inducing contraction of the smooth muscle surrounding the EOR to promote fluid movement towards the ovary. Our E16.5 ScRNA-seq data indicates that the EOR only expresses Arl13b and does not express the multicilia marker Foxj1. However, both markers are present in the putative EOR cell clusters in our adult SnRNA-seq data (Anbarci et al., 2024) suggesting that, after maturation, the EOR likely includes Arl13b+/Foxj1+ multiciliated cells. It is well documented that adult oviductal multiciliated cells express both ARL13b and FOXJ1 (Coy et al., 2016). It is possible that the primary cilia give rise to motile multiciliated cells, as shown in the airway epithelium (Jain et al., 2010) and suggested in the oviduct (Shi et al., 2014).

We used a “milking” process to extrude the contents of EOR, and mass spectrometry, to identify 4,232 proteins. We expected this group to be a mix of proteins from the luminal fluid, and cytoplasmic proteins from lysed EOR cells or from cells closely associated with the EOR. We took a very conservative approach to identify proteins highly likely to be secreted from EOR cells: (1) we compared the proteomics data with the known mammalian secretome to identify only secreted proteins; and (2) we next compared this shortened list with our available E16.5 ScRNA-seq data to find proteins whose transcripts were specific to the EOR at that stage. Although we recognize that the secretome at P7 is likely far more complex, this candidate list of 15 proteins reveals the first insight into specific proteins secreted from the RO that may be involved in the secretory process itself or may be transported to the ovary to affect ovary function.

Within the protein candidate list, we focused on validating clusterin (CLU) and syntaxin binding protein 2 (STXBP2) due to their roles in protein and vesicle transport. Consistent with our proteomic results, both Clu and Stxbp2 RNAs were detected in the RO. CLU is a chaperone protein that is present in both reproductive and non-reproductive tissues (Argraves and Morales, 2004; Saewu et al., 2017). In the epididymis, CLU acts as a chaperone to direct luminal proteins to the sperm head surface (Saewu et al., 2017). Although the role of CLU in the RO remains unknown, we hypothesize that it is important for post-secretory protein guidance to the ovary. STXBP2 binds t-SNARE protein STX and is essential for the vesicle-apical surface membrane fusion stage of vesicle secretion (Söllner, 2003). The presence of STXBP2 in the candidate list, coupled with validation of the expression of Stxbp2 RNA in the cells of the EOR, led us to investigate the presence of other components of the SNARE complex in the EOR.

The SNARE complex is comprised of a vesicle bound v-SNARE (Vamps), membrane bound t-SNARE (STXs), small GTPases (RABs), and syntaxin binding proteins (STXBPs) (Söllner, 2003). Using IF staining, we confirmed the presence of STX3, as well as RAB11. Our ScRNA-seq data indicate that EOR cells also express Vamp7 and Vamp8, which are v-SNAREs typically associated with STX3 (Dingjan et al., 2018). Small GTPases, such as RAB11, are essential in the SNARE complex as they mediate the tethering of vesicle membranes to the apical membrane prior to membrane fusion. Taking into consideration the presence of all components of the SNARE complex, we hypothesize that the EOR utilizes the SNARE complex to promote apical secretion into the luminal fluid.

A protein that was prominent on the proteomic candidate list was IGFBP2. We found that Igfbp2 was highly expressed in both the embryonic and adult RO transcriptome (Anbarci et al., 2024). IGFBP2 is an ideal candidate protein to mediate ovarian function. The IGF1 pathway is essential for regulation of follicular growth and selection (Talia et al., 2021). IGFBP2 can bind and sequester IGF1 (Rosenzweig, 2004), limiting its availability to promote follicle growth through activation of its receptor IGF1R (Amutha and Rajkumar, 2017; Bezerra et al., 2018; Liu et al., 2021). We spatially validated the expression of Igfbp2 in the EOR using HCR, and noted high expression at E18.5, which was decreased by P7. This decrease correlated with the end of the first-wave of follicle activation immediately after birth (Mork et al., 2012; Zheng et al., 2014). We also found variation in levels of Igfbp2 in our adult SnRNA-seq data, where Igfpb2 was enriched specifically in RO cells during the estrus stage compared to the diestrus stage of the estrous cycle. Based on these results, we hypothesize that high levels of estradiol during estrus trigger the EOR to produce IGFBP2 to sequester free IGF1 and decrease follicle growth and production of estradiol, maintaining ovary homeostasis.

The EOR is highly integrated with the extra-ovarian environment. We found that EOR cells are tightly surrounded in a dense web of vasculature. The RO may send or receive information through the vasculature, similar to other epithelial tissues in the female reproductive tract (Reynolds et al., 2002), raising the possibility that the RO participates in endocrine signaling. We also found that the EOR is directly contacted by TUJ1+ neurons, another avenue through which information may be entering or exiting the EOR. It is unclear whether the innervation contacts the epithelial cells specifically, similar to the way in which innervation contacts epithelial cells in the gut (Bohórquez et al., 2015), or whether the innervation contacts the smooth muscle cells that surround the EOR to induce contraction, similar to the innervation-muscle interaction in the uterus (Morizaki et al., 1989). We also found that the EOR is specifically associated with F4/80+ LYVE1 - macrophages. LYVE1 macrophages are predicted to have an angiogenic role in the adipose tissue surrounding the epididymis and could possibly be playing a similar role in the EOR (Cho et al., 2007). The integration of the EOR with the extra-ovarian environment, including vasculature, neurons, and macrophages, places the EOR in an ideal position to act as an antenna to interpret homeostatic signals and send information to the ovary. In Drosophila and C. Elegans, there is ample evidence that the ovary responds to physiologic levels of glycogen, insulin, amino acid levels, and changes in diet (Hubbard et al., 2013; Laws and Drummond-Barbosa, 2017). The mediator of this response has not been determined in mammals, but we hypothesize that this is the function of the RO. Interestingly, our scRNAseq data revealed expression of ESR1, PGR, INSR, and IGF1R, which are all critical hormone receptors regulating different aspects of female reproduction and health. Ongoing work will investigate how the EOR responds to hormones and other physiological signals, and whether proteins secreted by the EOR such as IGFBP2 respond to physiologic stimuli such as diet or immune status, and convey this information to the ovary.

Methods and Materials

Key Resource Table

Mice and Tissue Collection

Unless otherwise stated, mice used for experiments were maintained on the CD-1 or mixed CD-1 and C57BL/6J genetic backgrounds. The Pax8-rtTA and Tre-H2BGFP lines were previously described (Traykova-Brauch et al., 2008; Tumbar et al., 2004) and maintained on a mixed CD-1/C57BL/6J background, and maintained on a mixed CD-1/C57BL/6J background. The Pax8-rtTa;Tre-Cre;mTmG line was obtained by crossing the Pax8-rtTa line with carriers of the Tre-Cre (Perl et al., 2002) and mTmG (Muzumdar et al., 2007) alleles, which allowed for visualization of cell membranes of PAX8+ cells. Pregnant and nursing dams were given a doxycycline diet at 625 mg/kg (Teklad Envigo TD.01306) 3 days prior to tissue collection to induce GFP expression in Pax8+ cells. Toe samples were collected from mice for genotyping. Primers used for PCR genotyping are listed in the supplementary table 1. To obtain samples at specific stages of development, males were housed with females for timed matings. Successful mating was determined by the presence of a vaginal plug. Date of plug was considered embryonic day 0.5. Tissue samples were collected in phosphate-buffered saline without calcium or magnesium (PBS -/-), fixed in 4% paraformaldehyde (PFA)/PBS for 30 minutes at room temperature and dehydrated stepwise into 100% methanol, followed by storage at −20°C. All mice were housed in accordance with National Institutes of Health guidelines, and experiments were conducted with the approval of the Duke University Medical Center Institutional Animal Care and Use Committee (protocol #: A078-23-03).

Immunostaining and Confocal Image Acquisition

Samples were stepwise rehydrated into 100% PBS, followed by a 30-minute permeabilization wash in PBS 0.1% Triton X-100. Samples were then blocked for 1 hour with PBS, 1% Triton X-100, 10% horse serum, and 3% BSA. Samples were incubated overnight in primary antibodies diluted in blocking solution at 4°C (key resource table). Samples then underwent three 30 - minute washes in permeabilization solution and incubated overnight in secondary antibodies (1:500 dilution) and Hoechst vital dye solution diluted in blocking solution. The next day, samples underwent three 20-minute washes in permeabilization solution before being transferred into 100% PBS and stored at 4°C until ready for confocal imaging. Samples were mounted in DABCO mounting solution and stored at −20°C until imaged. Samples were imaged from both the dorsal and ventral sides using 3D-printed reversible slides that utilize two coverslips that allow for flipping. The 3D model can be downloaded on the NIH 3D Print Exchange website at https://3dprint.nih.gov/discover/3DPX-009765. Samples were imaged in toto using laser scanning confocal microscopy captures in the longitudinal plane on Zeiss LSM780 or LSM880 and affiliated Zen software Carl Zeiss, Inc, Germany) using 10x, 20x, and 63x (also used for Airyscan) objectives.

Hybridization chain reaction

The mouse embryo protocol from Molecular Instruments whole-mount (https://files.molecularinstruments.com/MI-Protocol-RNAFISH-Mouse-Rev9.pdf) was adjusted for P0 and P7 ovary samples. Samples were stepwise rehydrated into 0.1% Tween 20 / PBS (PBST). Samples were then subjected to 10 μg/mL proteinase K solution for 10 minutes at room temperature, then washed twice in PBST for 5 minutes. Samples were then post-fixed in 4% PFA for 10 minutes at room temperature, followed by three PBST washes for 5 minutes. Samples were then pre-hybridized for 30 minutes in 500ul of hybridization buffer (Molecular Instruments) at 37°C. Samples were then incubated overnight at 37°C with HCR probes diluted in 500ul hybridization buffer (2pmol in 500ul). Samples were then washed for 15 minutes four times in the Molecular Instruments probe wash buffer at 37°C, followed by two 5x 0.1% Tween 20/SSC (SSCT) washes for 5 minutes at room temperature. Fluorescently labeled hairpins were snap cooled and left in a dark drawer at room temperature for 30 minutes. Following hairpin preparation, samples were incubated with hairpins diluted in Molecular Instruments amplification buffer overnight in the dark at room temperature. On the third day, samples were washed in SSCT four times for 15 minutes each before being transferred to PBS -/- and then mounted in DABCO mounting medium for imaging.

Image Processing

Confocal images were imported into FIJI software for minor image processing (cropping, rotations, maximum intensity projection/ optical slice montage, and color application). The RO was oriented to the left of the ovary with the oviduct at the top of the ovary. Images were then imported into Adobe Photoshop CC (Adobe, Inc, CA) for final processing of channel overlay, brightness, contrast adjustment, and modification of red to “magenta” (hue adjustment to −30). Channel color accessibility was determined via Photoshop colorblind proofing.

Dextran Injections

Postnatal day 7 mice were euthanized, and the entire ovarian complex was dissected in PBS +/+. Samples were then placed on agar blocks (1.5%) soaked in PBS +/+. The distal tip of the EOR was identified and punctured using a tungsten needle (0.001mm tip diameter, Fine Science Tools, 10130 −05). A microinjection unit (Picospritzer III microinjector) and fine capillary glass needle (5-10 um) loaded with 2-3 nl dextran solution (1.25mg/ml dextran, PBS, 0.05% phenol red) was used to inject dextran into the opening of the EOR. Samples were then left for 15 minutes before fixing with 4% PFA. During the 15 minutes, samples were monitored to visualize movement. Samples were then fixed, stained, and imaged.

Luminal Fluid Collection for Proteomics and Analysis

Postnatal day 7 EORs were collected from Pax8-rtTa; Tre-H2B-Gfp mice (as a guide to only collect EOR) and carefully cleaned up to remove as much non-EOR tissue (40 EOR samples). Samples were then placed in a 1.5ml Eppendorf tube and ‘pressed’ for fluid using a disposable pellet pestle. The sample was then spun down and the supernatant was collected and snap frozen in liquid nitrogen. The sample was then submitted to the Duke University Proteomics Core for mass spectrometry. Proteomic analysis uncovered 4,252 proteins present in the pressed fluid. To exclude proteins that may have contributed due to cell lysis, results were cross-referenced against a database of secreted proteins (Meinken et al., 2015). Overlapping proteins were then cross-referenced to our E16.5 ScRNA-seq data (Anbarci et al., 2024), and candidate proteins were determined by gene expression specific to the EOR cluster.

LC-MS/MS Proteomics Analysis

The sample was subjected to a Bradford (Pierce) protein measurement and 10 ug was removed for downstream processing. The sample was brought to 4% SDS, reduced with 10 mM dithiothreitol for 20 min at 55°C, alkylated with 25mM iodoacetamide for 45 min at room temperature and then subjected to S-trap (Protifi) trypsin digestion using manufacturer recommended protocols. Digested peptides were lyophilized to dryness and resuspended in 50 uL of 0.2% formic acid/2% acetonitrile. The sample was subjected to chromatographic separation on a Waters MClass UPLC equipped with a 1.8 μm Acquity HSS T3 C18 75 μm × 250 mm column (Waters Corp.) with a 90-min linear gradient of 5 to 30% acetonitrile with 0.1% formic acid at a flow rate of 400 nanoliters/minute (nL/min) with a column temperature of 55°C. Data collection on the Fusion Lumos mass spectrometer with a FAIMS Pro device was performed for three difference compensation voltages (−40v, −60v, −80v). Within each CV, a data-dependent acquisition (DDA) mode of acquisition with a r=120,000 (@ m/z 200) full MS scan from m/z 375 – 1500 with a target AGC value of 4e5 ions was performed. MS/MS scans with HCD settings of 30% were acquired in the linear ion trap in “rapid” mode with a target AGC value of 1e4 and max fill time of 35 ms. The total cycle time for each CV was 0.66s, with total cycle times of 2 sec between like full MS scans. A 20s dynamic exclusion was employed to increase depth of coverage. The total analysis cycle time for each sample injection was approximately 2 hours.

Raw LC-MS/MS data files were processed in Proteome Discoverer 3.0 (Thermo Scientific) and then submitted to independent Sequest database searches against a Mus musculus protein database containing both forward and reverse entries of each protein. Search tolerances were 2 ppm for precursor ions and 0.8 Da for product ions using trypsin specificity with up to two missed cleavages. Carbamidomethylation (+57.0214 Da on C) was set as a fixed modification, whereas oxidation (+15.9949 Da on M) was considered a dynamic mass modifications. All searched spectra were imported into Scaffold (v5.3, Proteome Software) and scoring thresholds were set to achieve a peptide false discovery rate of 1% using the PeptideProphet algorithm. Data were output as total spectral matches.